Meta-analysis of gut microbiome studies identifies disease-specific and shared responses

### 2018-12-10 meta-analysis/manuscript.tex
Meta-analysis ms + supp, and fix figures and tables

\section*{Introduction}

@@

\section*{Results}

\subsection*{Most disease states show altered microbiomes}

@@

\subsection*{Loss of beneficial microbes or enrichment of pathogens}

@@

\subsection*{Shared vs. disease-specific microbial responses}

@@

\subsection*{Within and cross-disease meta-analysis improves interpretability}

@@

\section*{Discussion}

@@

\section*{Methods}

\subsection*{Dataset collection}
We identified case-control 16S studies from keyword searches in PubMed and by following references in meta-analyses and related case-control studies.
We included studies with publicly available raw 16S data (fastq or fasta) and metadata indicating case or control status for each sample.
Most data was downloaded from online repositories (e.g. SRA) or links provided in the original publications, but some were acquired after personal communication with the authors (Supplementary Table \ref{tab:data}).
We did not include any studies which required additional ethics committee approvals or authorizations for access (e.g. controlled dbGaP studies).
In studies where multiple body sites were sampled or where multiple samples were taken per patient, we also required the respective metadata to include those metadata.
We analyzed only stool 16S samples, and excluded studies with fewer than 15 case patients.
In CRC studies with multiple control groups (e.g. healthy and non-CRC adenoma), only the healthy patients were used as controls for all of our comparisons.
In studies with non-healthy controls (e.g. non-IBD patients), these patients were used as controls (as in the original papers).
In the Schubert et al. CDI study \cite{cdi-schubert}, which had both CDI and non-CDI diarrheal patients, each group was used as an independent case group compared with controls.
We also analyzed the NASH and obese patients from the Zhu et al. study \cite{nash-baker} as independent case groups.
When obesity studies reported body mass index instead of obesity status, we considered patients with BMI less than 25 as our control group and patients with BMI greater than 30 as the case group.

@@

\subsection*{Statistical analyses}
To perform supervised classification of cases and controls within each dataset, we built Random Forest classifiers with 5-fold cross-validation.
To build our train and test sets, we used the python \texttt{scikit-learn} \texttt{StratifiedKFold} function with shuffling of the data \cite{scikit-learn}.
To build our classifiers, we used the \texttt{RandomForestClassifier} function with 1000 estimators and other default settings \cite{scikit-learn}.
We found no significant effect of various Random Forest parameters on the AUCs  (Supplementary Figures \ref{fig:rf_params_gini} and \ref{fig:rf_params_entropy}).
We calculated the interpolated area under the ROC curve (AUC) for each classifier based on the cross-validation testing results.
To account for spurious high classifiability due to class imbalances, we also calculated the Cohen's kappa score for each classifier using \texttt{sklearn.metrics.cohen\_kappa\_score} on the test set predictions (Supplementary Table \ref{tab:kappa}).
The kappa scores correlated well with the AUCs (Pearson $\rho$ $=$ 0.9), indicating that the majority of the classifiers performed well even when considering their underlying data distributions.
We excluded Youngster et al. (2014) \cite{cdi-youngster}, which had only 4 distinct control patients, from all classifier analyses.

@@

\subsection*{Microbiome community analyses}

@@

\subsection*{Code availability}

@@

\subsection*{Data availability}

@@

\section*{Acknowledgements}

@@

\section*{Author contributions}

@@

\section*{Competing Financial Interests}

@@

@@

\newpage
\begin{figure}[h]
	\begin{center}
  	\makebox[\textwidth][c]{\includegraphics[width=1.2\textwidth]{samplesize_auc_nsig_balance.pdf}}
	\caption{\textbf{Most diseases show microbiome alterations, and consistent disease-associated shifts differ in their extent and direction.} (A) Left: Total sample size for each study included in these analyses. Additional information about each dataset can be found in Table \ref{tab:datasets}. Studies on the y-axis are grouped by disease and ordered by decreasing sample size (top to bottom). Right: Area under the ROC curve (AUC) for genus-level random forest classifiers. X-axis starts at 0.5, the expected value for a classifier which assigns labels randomly, and AUCs less than 0.5 are not shown. ROC curves for all datasets are in Supplementary Figure \ref{fig:roc_curves}. Note that Youngster et al. (2014) \cite{cdi-youngster} had only 4 distinct control patients was excluded from the Random Forest analysis. (B) Left: Number of genera with q $<$ 0.05 (Kruskal-Wallis (KW) test, Benjamini-Hochberg FDR correction) for each dataset. If a study has no significant associations, no point is shown. Right: Direction of the microbiome shift, i.e. the percent of total associated genera which were enriched in diseased patients. In datasets on the leftmost blue line, 100\% of associated (q $<$ 0.05, FDR KW test) genera are health-associated (i.e. depleted in patients relative to controls). In datasets on the rightmost red line, 100\% of associated (q $<$ 0.05, FDR KW test) genera are disease-associated (i.e. enriched in patients relative to controls). Supplementary Figures \ref{fig:overall_heatmap_qvalues} and \ref{fig:overall_heatmap_foldchange} show q-values and effects for each genus in each study.}
	\label{fig:fig1}
	\end{center}
\end{figure}

\newpage
\begin{figure}[h]
	\begin{center}
  	\makebox[\textwidth][c]{\includegraphics[width=0.9\textwidth]{disease_specific_heatmaps.pdf}}
	\caption{\textbf{Comparing results from multiple studies of the same disease reveals patterns in disease-associated microbiome alterations.} Heatmaps showing log10(q-values) for each disease (Kruskal-Wallis (KW) test, Benjamini-Hochberg FDR correction). Rows include all genera which were significant in at least one dataset within each disease, columns are datasets. Q-values are colored by direction of the effect, where red indicates higher mean abundance in disease patients and blue indicates higher mean abundance in controls. Opacity ranges from q = 0.05 to 1, where q values less than 0.05 are the most opaque and q values close to 1 are gray. White indicates that the genus was not present in that dataset. Within each heatmap, rows are ordered from most disease-associated (top) to most health-associated (bottom) (i.e. by the sum across rows of the log10(q-values), signed according to directionality of the effect). The extent of a disease-associated microbiome shift can be visualized by the number of rows in each disease heatmap; the directionality of a shift can be seen in the ratio of red rows to blue rows within each disease. See Supplementary Figure \ref{fig:supp_dis_specific} for genus (row) labels.}
	\label{fig:fig2}
	\end{center}
\end{figure}

\newpage
\begin{figure}[h]
	\begin{center}
  	\makebox[\textwidth][c]{\includegraphics[width=1.2\textwidth]{core_response.pdf}}
	\caption{\textbf{The majority of disease-associated microbiome alterations overlap with a non-specific microbial response to disease.} (A) Non-specific and disease-associated genera. Genera are in columns, arranged phylogenetically according to a PhyloT tree built from genus-level NCBI IDs (\url{http://phylot.biobyte.de}). Non-specific genera are associated with health (or disease) in at least two different \textit{diseases} (q $<$ 0.05, Kruskal-Wallis (KW) test, Benjamini-Hochberg FDR correction). Disease-specific genera are significant in the same direction in at least two \textit{studies} of the same disease (q $<$ 0.05, FDR KW test). As in Figure \ref{fig:fig2}, blue indicates higher mean abundance in controls and red indicates higher mean abundance in patients. Black bars indicate mixed genera which were associated with health in two diseases and also associated with disease in two diseases. Disease-specific genera are shown for diseases with at least 3 studies. Phyla, left to right: Euryarchaeota (brown), Verrucomicrobia Subdivision 5 (gray), Candidatus Saccharibacteria (gray), Bacteroidetes (blue), Proteobacteria (red), Synergistetes (pink), Actinobacteria (green), Firmicutes (purple), Verrucomicrobia (gray), Lentisphaerae (pink), Fusobacteria (orange). See Supplementary Figure \ref{fig:supp_meta_heatmap} for genus labels. (B) The percent of each study's genus-level associations which overlap with the shared response  (q $<$ 0.05, FDR KW test). Only datasets with at least one significant association are shown. (C) Overall abundance and ubiquity of non-specific genera across all patients in all datasets. Non-specific genera on the x-axis are as defined above.}
	\label{fig:fig3}
	\end{center}
\end{figure}

\renewcommand{\figurename}{Supplementary Figure}
\renewcommand{\tablename}{Supplementary Table}
\setcounter{figure}{0}
\setcounter{table}{0}

\section*{Supplementary Note 1: Re-processing and re-analyzing raw data yields results which are generally consistent with previously published results}\label{sec:lit_comp}

Our re-analyses of the 29 studies were largely consistent with the originally reported results, with the same taxonomic groups showing similar trends despite differences in data-processing methodologies.
We usually found fewer significant (q $<$ 0.05) differences between control and diseased groups, which is likely due to our choice of a non-parametric statistical test (Kruskall-Wallis) paired with a multi-test correction (FDR).
Thus, our results are more conservative.
We also collapsed to genus level in order to compare results across disparate studies, which prevented us from identifying species- or strain-specific associations which the original authors may have identified.
A major advantage of our re-analysis is that each data set was processed and analyzed in the same way, which allowed us to more directly compare results across studies and diseases.

\subsection*{\textit{Clostridium difficile} Infection and enteric diarrhea are characterized by large-scale shifts in the microbiome (CDI; 4 studies)}

Schubert et al. (2014) looked at how the gut microbiota differed between CDI patients with diarrhea (n $=$ 94), non-CDI patients with diarrhea (n $=$ 89), and non-diarrheal controls (n $=$ 155) \cite{cdi-schubert}.
Similar to other CDI studies, the authors found a significant reduction in alpha diversity in patients with diarrhea (Dunn’s multiple-comparison test on AMOVA, p $<$ 0.0001).
They found that OTUs from the \textit{Ruminococcaceae}, \textit{Lachnospiraceae}, \textit{Bacteroides}, \textit{Prevotellaceae}, and \textit{Porphyromonadaceae} families were enriched in healthy subjects relative to patients with CDI and non-CDI diarrhea.
They also showed that OTUs from the \textit{Enterococcus} genus and the \textit{Enterobacteriaceae} and \textit{Erysipelotrichaceae} families were more prevalent in patients with diarrhea.
In our analysis of the data, we also observed a significant reduction in alpha diversity in patients with diarrhea (q $<=$ 0.05, KW test).
Similarly, we found that \textit{Enterobacteriaceae}, \textit{Enterococcus}, and \textit{Erysipelotrichaceae} were enriched in CDI patients, in addition to \textit{Fusobacterium}, \textit{Parvimonas}, \textit{Veillonella}, \textit{Carnobacterium}, \textit{Streptococcus}, \textit{Tetragenococcus}, \textit{Lactobacillus}, \textit{Pediococcus}, \textit{Gemella}, \textit{Staphylococcus}, \textit{Butyricicoccus}, \textit{Robinsoniella}, \textit{Clostridium XlVa}, \textit{Clostridium XlVb}, \textit{Ruminococcus2}, \textit{Flavonifractor}, \textit{Gemmiger}, \textit{Mogibacterium}, \textit{Peptostreptococcus}, \textit{Clostridium XI}, \textit{Eggerthella}, \textit{Atopobium}, \textit{Actinomyces}, \textit{Arthrobacter}, \textit{Aggregatibacter}, \textit{Pseudomonas}, and \textit{Dysgonomonas}.
As in the original study, we found that \textit{Bacteroides}, \textit{Alistipes}, \textit{Anaerovorax}, \textit{Oxalobacter}, \textit{Bordetella}, \textit{Prevotellaceae}, \textit{Porphyromonadaceae}, \textit{Lachnospiraceae}, and \textit{Ruminococcaceae} were more abundant in the healthy controls.
We also found \textit{Turicibacter}, \textit{Dialister}, \textit{Eubacterium}, \textit{Asteroleplasma}, \textit{Cloacibacillus}, \textit{Bordetella}, \textit{Oxalobacter}, \textit{Sutterella}, \textit{Parasutterella}, \textit{Desulfovibrio}, \textit{Sediminibacterium}, and \textit{Methanobrevibacter} to be enriched in the controls (q $<=$ 0.05, KW tests).
Overall, our analysis closely matched what was presented in the original manuscript.

Vincent et al. (2013) compared 25 patients with CDI to 25 healthy control patients \cite{cdi-vincent}.
The authors found a significant reduction in alpha diversity (p $<=$ 0.05, Mann-Whitney U test).
They also report a reduction in \textit{Bacteroidaceae} and \textit{Clostridiales Incertae Sedis XI} in CDI patients relative to controls, and an enrichment in \textit{Enterococcaceae} in CDI patients (p $<$ 0.05, logistic regression).
After reprocessing these data and collapsing abundances to the genus level, we observed a similar reduction in alpha diversity (q $<=$ 0.05, KW test). We saw that the \textit{Enterococcaceae} genera \textit{Enterococcus} and \textit{Proteus} were enriched in CDI patients. Healthy controls showed higher levels of \textit{Fusobacterium}, \textit{Peptoniphilus}, \textit{Murdochiella}, \textit{Anaerococcus}, \textit{Finegoldia}, \textit{Odoribacter}, \textit{Prevotella}, and \textit{Parabacteroides} relative to CDI patients. In summary, our results are fairly similar to the authors' original analysis, showing a depletion in \textit{Bacteroidetes} and an enrichment in \textit{Proteobacteria} in CDI patients.

Youngster et al. (2014) applied fecal microbiota transplants (FMTs) with materials collected from 5 healthy donors to 20 patients with recurrent \textit{Clostridium difficile} infections (CDIs) \cite{cdi-youngster}.
The goal of this study was to determine whether nasal-gastric tube or colonoscopy administration of FMTs was most effective for treating CDIs (i.e. half of the CDI patients received one or the other treatment).
The authors reported a significant reduction in alpha diversity in CDI patients vs. the healthy donors (p $<$ 0.001, Mann-Whitney test).
They did not assess whether there were significant differences in microbial community composition between CDI patients and donors, although they show that composition becomes more similar to donors following FMT.
In our analysis, we also found a significant reduction in alpha diversity (p $<=$ 0.05, KW test). \textit{Enterococcus} was enriched in CDI patients relative to healthy stool donors (q $<=$ 0.05, KW tests) and 15 genera were depleted in CDI patients relative to healthy controls.
Healthy donors were enriched in genera from \textit{Ruminococcaceae} and \textit{Lachnospiraceae} families, in addition to the genera \textit{Dialister} and \textit{Anaerosporobacter}.

Singh et al. (2015) examined differences in the gut microbiome between individuals with enteric infections (n=200) and healthy controls (n=75) \cite{edd-singh}.
The authors report a significant drop in alpha diversity in diseased patients relative to the controls (unknown test).
They also report a general reduction in the dominance of \textit{Firmicutes} and \textit{Bacteroidetes} phyla and an increase in the prevalence of \textit{Proteobacteria} in diseased patients.
Specifically, they report an increase in the abundance of \textit{Enterobacteriaceae}, \textit{Lactobacillaceae}, \textit{Pasteurellaceae}, \textit{Streptococcus}, \textit{Bacilli}, \textit{Escherichia},  \textit{Haemophilus}, and certain \textit{Ruminococcus} species in patients with diarrhea.
In healthy people, they report a significant enrichment in \textit{Verrucomicrobia}, \textit{Dorea}, \textit{Blautia}, \textit{Holdermania}, \textit{Ruminococcaceae}, \textit{Lachnospiraceae}, \textit{Butyricimonas}, \textit{Faecalibacterium}, \textit{Bacteroidaceae}, and \textit{Bifidobacterium}, \textit{Sutterella}, \textit{Parabacteroides}, \textit{Rikenellaceae}, and \textit{Oscillospira}.
After re-processing the data, we found very similar results to those originally reported.
We found that alpha diversity was significantly lower in patients with enteric infections (q $<=$ 0.05, KW test).
We saw significant enrichment in \textit{Proteobacteria} families in patients with diarrhea, including \textit{Enterobacteriaceae}, \textit{Pasteurellaceae}, \textit{Campylobacteraceae}, and \textit{Neisseriaceae}.
We also saw higher levels of \textit{Fusobacterium}, \textit{Parvimonas}, \textit{Veillonella}, \textit{Lactococcus}, \textit{Streptococcus}, \textit{Enterococcus}, \textit{Tetragenococcus}, \textit{Gemella}, \textit{Ruminococcus II}, \textit{Peptostreptococcus}, and \textit{Collinsella} in diseased patients.
In the healthy controls, we found enrichment of 43 genera, including \textit{Sutterella}, \textit{Verrucomicrobia} (\textit{Akkermansia}), \textit{Ruminococcaceae}, \textit{Lachnospiraceae}, \textit{Bacteroidaceae}, and \textit{Bifidobacterium}.
In addition, we saw higher levels of several members of \textit{Rumminococcaceae}, \textit{Lachnospiraceae}, and \textit{Bacteroidales} in healthy controls (q $<=$ 0.05, KW tests).
Overall, our results largely overlap with those presented, but we identify a number of significant taxa that were not originally reported.

Taken together, we see large-scale shifts in the microbiome associated with both CDI and non-CDI diarrhea.
The dysbiosis of enteric infection and diarrhea is quite consistent across studies.
In general, \textit{Proteobacteria} increase in prevalence in patients with diarrhea, with a concomitant decrease in \textit{Bacteroidetes} and \textit{Firmicutes}.
In particular, we see a reduction in butyrate-producing Clostridia, including genera within \textit{Ruminococcaceae} and \textit{Lachnospiraceae} families, which have been associated with a healthy gut.
We also see in increase in prevalence of organisms often associated with lower pH and higher oxygen levels of the upper-gut, like \textit{Lactobacillaceae} and \textit{Enterobacteriaceae} \cite{donaldson2016gut}, in patients with diarrhea.
Thus, diarrhea leads to consistent and large-scale rearrangements in the composition of the gut microbiome.

\subsection*{Colorectal Cancer has a consistent, ootentially pathogenic microbial signature (CRC; 4 studies)}

Baxter et al. (2016) looked at differences in the microbiomes of 120 colorectal cancer (CRC) patients, 198 patients with non-cancerous adenomas, and 172 healthy controls \cite{crc-baxter}.
Similar to prior work, the authors found that \textit{Porphyromonas}, \textit{Peptostreptococcus}, \textit{Parvimonas}, and \textit{Fusobacterium} were positively associated with CRC (random forest classifiers).
Furthermore, they found that the absence of certain \textit{Lachnospiraceae} species was associated with the presence of adenomas.
We found similar patterns in our re-analysis of these data, with \textit{Fusobacterium}, \textit{Peptostreptococcus}, \textit{Parvimonas}, and \textit{Porphyromonas} enriched in CRC patients (q $<=$ 0.05, KW tests).
We also found higher levels of \textit{Victivallis}, \textit{Peptoniphilus}, \textit{Anaerococcus}, \textit{Catenibacterium}, \textit{Staphylococcus}, \textit{Collinsella}, \textit{Enterobacter}, and \textit{Alloprevotella} in CRC patients (q $<=$ 0.05, KW tests).
We found that healthy controls were enriched in \textit{Lachnobacterium} (genus within \textit{Lachnospiraceae}), \textit{Gemmiger} (within \textit{Rumminococcaceae}), \textit{Clostridium XVIII}, and \textit{Haemophilus} (q $<=$ 0.05, KW tests).
Overall, these results match what has been reported previously for CRC \cite{brennan2016gut}.

Zeller et al. (2014) collected microbiome data from 41 CRC patients and 75 control patients \cite{crc-zeller}.
At the phylum level, they found that \textit{Proteobacteria}, \textit{Fusobacteria}, and \textit{Bacteroidetes}, were more abundant in CRC patients, while \textit{Firmicutes} and \textit{Actinobacteria} were enriched in control patients.
At the genus level, the authors report higher levels of \textit{Fusobacterium}, \textit{Pseudoflavonifractor}, \textit{Peptostreptococcus}, \textit{Leptotrichia}, \textit{Porphyromonas}, \textit{Desulfovibrio}, \textit{Parvimonas}, \textit{Selenomonas}, and \textit{Bilophila} in CRC patients (q $<=$ 0.1, FDR-corrected Wilcoxon tests).
Healthy controls were enriched in \textit{Bifidobacterium}, \textit{Acinetobacter}, \textit{Campylobacter}, \textit{Ruminococcus}, and \textit{Eubacterium} genera (q $<=$ 0.1, FDR-corrected Wilcoxon tests).
In our re-analysis we found enrichment of \textit{Fusobacterium}, \textit{Parvimonas}, \textit{Flavonifractor}, \textit{Anaerotruncus}, \textit{Anaerovorax}, \textit{Peptostreptococcus}, \textit{Comamonas}, \textit{Eikenella}, \textit{Butyricimonas}, and \textit{Porphyromonas} genera in CRC patients (q $<=$ 0.05, KW tests).
In healthy patients, we found higher levels of \textit{Anaerostipes} (within \textit{Lachnospiraceae}; q $<=$ 0.05, KW tests).

Wang et al. (2011) analyzed a cohort of 46 CRC patients and 56 healthy controls \cite{crc-zhao}.
The authors found no difference in alpha diversity between CRC and control patients.
CRC patients had higher abundances of \textit{Porphyromonas}, \textit{Escherichia-Shigella}, \textit{Enterococcus}, \textit{Streptococcus}, and \textit{Peptostreptococcus} genera (p $<=$ 0.05, Mann-Whitney).
The authors report that healthy controls were enriched \textit{Bacteroides}, \textit{Roseburia}, \textit{Alistipes}, \textit{Eubacterium}, and \textit{Parasutterella} genera (p $<=$ 0.05, Mann-Whitney).
We found very similar results in our re-analysis of these data.
We saw greater levels of \textit{Enterococcus}, \textit{Peptostreptococcus}, \textit{Enterobacter}, \textit{Klebsiella}, \textit{Escherichia-Shigella}, and \textit{Porphyromonas} genera in CRC patients (q $<=$ 0.05, KW tests).
And we observed significantly higher levels of \textit{Bacteroides}, and several genera within \textit{Lachnospiraceae} in healthy controls (q $<=$ 0.05, KW tests).
Furthermore, we also did not detect any significant differences in alpha diversity between CRC and healthy patients.

Chen et al. (2012) analyzed stool from 22 healthy patients and 21 CRC patients \cite{crc-xiang}.
The authors found that \textit{Paraprevotella}, \textit{Eubacterium}, \textit{Desulfovibrio}, \textit{Mogibacterium}, \textit{Collinsella}, \textit{Anaerotruncus}, \textit{Slackia}, \textit{Anaerococcus}, \textit{Porphyromonas}, \textit{Fusobacterium}, and \textit{Peptostreptococcus} genera were significantly enriched in CRC patients relative to controls, while \textit{Bifidobacterium}, \textit{Faecalibacterium}, and \textit{Blautia} were reduced in CRC patients (p $<=$ 0.05, Mann-Whitney).
In our re-analysis of this data set, we found no significant differences between CRC and control patients.
Again, this is likely due to the small number of replicates and our implementation of multiple-test corrections.
However, non-significant trends were largely in agreement with the original results.

Across these four colorectal cancer studies, we find significant agreement.
Dysbiosis associated with CRC is generally characterized by increased prevalence of \textit{Fusobacterium}, \textit{Porphyromonas}, \textit{Peptostreptococcus}, \textit{Parvimonas}, \textit{Leptotrichia}, \textit{Desulfovibrio}, and \textit{Anaerococcus} genera (i.e. these genera were higher in CRC patients in 2 or more studies).
In addition, there is a consistent decrease in the abundances of \textit{Faecalibacterium}, \textit{Blautia}, \textit{Bacteroides} genera and organisms from the \textit{Lachnospiraceae} family in CRC patients.
CRC appears to have a smaller impact on overall community structure than diahrrea.
Indeed, we saw no significant differences in alpha diversity between healthy controls and CRC patients.
In summary, CRC is characterized by a consistent enrichment of disease-associated bacteria.

\subsection*{Inflammatory Bowel Disease is characterized by a depletion of health-associated bacteria (IBD - Ulcerative Colitis and Crohn's Disease; 4 studies)}

Gevers et al. (2014) looked for microbial signatures of Crohn's disease (CD) samples across 447 CD patients and 221 non-IBD controls \cite{ibd-gevers}.
Non-IBD controls were patients with non-inflammatory conditions such as abdominal pain and diarrhea.
The authors report increased abundance of \textit{Enterobacteriaceae}, \textit{Pasteurellaceae}, \textit{Veillonellaceae}, and \textit{Fusobacteriaceae} in CD patients.
CD patients also showed a drop in the abundances of \textit{Erysipelotrichales}, \textit{Bacteroidales}, and \textit{Clostridiales} (\textit{Ruminococcaceae} and \textit{Lachnospiraceae}) taxa.
These results were based on a mixture of 16S amplicon and shotgun metagenomic sequencing.
In our re-analysis of the 16S stool data, we found significant enrichment in \textit{Anaerosporobacter}, \textit{Roseburia}, \textit{Hespellia}, \textit{Ruminococcus II}, \textit{Eubacterium}, \textit{Pseudoflavonifractor}, \textit{Sporobacter}, \textit{Ruminococcus}, \textit{Subdoligranulum}, \textit{Papillibacter}, \textit{Collinsella}, and \textit{Methanobrevibacter} in healthy patients (q $<=$ 0.05, KW tests).
The only genera that we saw significantly enriched in CD patients were \textit{Lactobacillus} and \textit{Acetanaerobacterium} (q $<=$ 0.05, KW tests).
We found a similar set of taxa enriched in the controls, but did not detect as many significant CD-enriched genera as the authors reported.
This is likely due to the fact that we restricted our analysis to the 16S stool data.
However, we saw non-significant trends in \textit{Enterobacteriaceae} and \textit{Veillonellaceae} consistent with the results reported in the original paper.

Morgan et al. (2012) studied a cohort of 119 CD patients, 74 UC patients, and 27 healthy controls \cite{ibd-hut}.
The authors found that healthy patients’ gut microbiomes were significantly enriched in \textit{Roseburia}, \textit{Phascolarctobacterium}, and an unclassified genus in the family \textit{Veillonellaceae} (multivariate linear model, q $<=$ 0.25).
Patients with UC showed significantly higher levels of \textit{Clostridiaceae} (multivariate linear model, q $<=$ 0.25).
In our re-analysis, we did not find any genera that were significantly enriched in IBD patients.
We found that healthy patients had significantly greater abundances of \textit{Ruminococcus}, and \textit{Gemmiger} relative to both UC and CD patients (q $<=$ 0.05, KW tests).
Additionally, CD patients were depleted in \textit{Clostridium IV} relative to healthy controls (q $<=$ 0.05, KW tests).

Papa et al. (2012) studied a cohort of 23 CD patients, 43 UC patients, and 24 non-IBD controls \cite{ibd-papa}.
Non-IBD controls were patients with symptoms such as: constipation, abdominal pain, gastroesophageal reflux, poor weight gain, diarrhea, blood in stool and oropharyngeal dysphagia.
At the genus level, they found that controls were enriched in \textit{Alistipes}, \textit{Subdoligranulum}, \textit{Anaerovorax}, \textit{Oscillibacter}, \textit{Parabacteroides}, \textit{Odoribacter}, \textit{Ruminococcus}, \textit{Butyricicoccus}, \textit{Akkermansia}, \textit{Anaerotruncus}, \textit{Sporobacter}, \textit{Phascolarctobacterium}, \textit{Lawsonia}, \textit{Ethanoligenens}, \textit{Peptococcus} relative to IBD patients (KW, q $<$ 0.01).
The only genus that was found to be enriched in IBD patients was \textit{Escherichia-Shigella}.
In our re-analysis, we also found \textit{Escherichia-Shigella} and \textit{Cronobacter} to be enriched in patients with IBD (q $<=$ 0.05, KW tests).
When comparing healthy controls with UC patients, we also found an enrichment of \textit{Haemophilus} in the UC patients.
Control patients showed higher abundances of \textit{Phascolarctobacterium}, \textit{Butyricicoccus}, \textit{Ruminococcus II}, \textit{Oscillibacter}, \textit{Ruminococcus}, \textit{Gemmiger}, \textit{Subdoligranulum}, \textit{Clostridium IV}, \textit{Odoribacter}, \textit{Alistipes}, and \textit{Parabacteroides} relative to all IBD patients (q $<=$ 0.05, KW tests).
Additionally, control patients were enriched in \textit{Clostridium XIVa}, \textit{Flavonifractor}, and \textit{Akkermansia} relative to UC patients.
Overall, our results match very closely what was found in the original paper.

>>>>

\section{Introduction}

@@

\section{Results}

\subsection{Most disease states show altered microbiomes}

@@

\subsection{Loss of beneficial microbes or enrichment of pathogens}

@@

\subsection{Shared vs. disease-specific microbial responses}

@@

\subsection{Within and cross-disease meta-analysis improves interpretability}

@@

\section{Discussion}

@@

\section{Methods}

\subsection{Dataset collection}

We identified case-control 16S studies from keyword searches in PubMed and by following references in meta-analyses and related case-control studies.
We included studies with publicly available raw 16S data (fastq or fasta) and metadata indicating case or control status for each sample.
Most data was downloaded from online repositories (e.g. SRA) or links provided in the original publications, but some were acquired after personal communication with the authors (Supplementary Table \ref{tab:data}).
We did not include any studies which required additional ethics committee approvals or authorizations for access (e.g. controlled dbGaP studies).
In studies where multiple body sites were sampled or where multiple samples were taken per patient, we also required the respective metadata to include those metadata.
We analyzed only stool 16S samples, and excluded studies with fewer than 15 case patients.
In CRC studies with multiple control groups (e.g. healthy and non-CRC adenoma), only the healthy patients were used as controls for all of our comparisons.
In studies with non-healthy controls (e.g. non-IBD patients), these patients were used as controls (as in the original papers).
In the Schubert et al. CDI study \cite{cdi-schubert}, which had both CDI and non-CDI diarrheal patients, each group was used as an independent case group compared with controls.
We also analyzed the NASH and obese patients from the Zhu et al. study \cite{nash-baker} as independent case groups.
When obesity studies reported body mass index instead of obesity status, we considered patients with BMI less than 25 as our control group and patients with BMI greater than 30 as the case group.

@@

\subsection{Statistical analyses}

To perform supervised classification of cases and controls within each dataset, we built Random Forest classifiers with 5-fold cross-validation.
To build our train and test sets, we used the python \texttt{scikit-learn} \texttt{StratifiedKFold} function with shuffling of the data \cite{scikit-learn}.
To build our classifiers, we used the \texttt{RandomForestClassifier} function with 1000 estimators and other default settings \cite{scikit-learn}.
We found no significant effect of various Random Forest parameters on the AUCs  (Supplementary Figures \ref{fig:rf_params_gini} and \ref{fig:rf_params_entropy}).
We calculated the interpolated area under the ROC curve (AUC) for each classifier based on the cross-validation testing results.
To account for spurious high classifiability due to class imbalances, we also calculated the Cohen's kappa score for each classifier using \texttt{sklearn.metrics.cohen\_kappa\_score} on the test set predictions (Supplementary Table \ref{tab:kappa}).
The kappa scores correlated well with the AUCs (Pearson $\rho$ $=$ 0.9), indicating that the majority of the classifiers performed well even when considering their underlying data distributions.
We excluded Youngster et al. (2014) \cite{cdi-youngster}, which had only 4 distinct control patients, from all classifier analyses.

@@

\subsection{Microbiome community analyses}

@@

\subsection{Code availability}

@@

\subsection{Data availability}

@@

\section{Acknowledgements}

@@

\section{Author contributions}

@@

\subsection{Competing Financial Interests}

@@

\section{Tables}

@@

\section{Figures}

\newpage
\begin{figure}[h]
	\begin{center}
  	\makebox[\textwidth][c]{\includegraphics[width=1\textwidth]{samplesize_auc_nsig_balance.pdf}}
    \captionsetup{font=footnotesize,labelfont=footnotesize}
	\caption{\textbf{Most diseases show microbiome alterations, and consistent disease-associated shifts differ in their extent and direction.} (A) Left: Total sample size for each study included in these analyses. Additional information about each dataset can be found in Table \ref{tab:datasets}. Studies on the y-axis are grouped by disease and ordered by decreasing sample size (top to bottom). Right: Area under the ROC curve (AUC) for genus-level random forest classifiers. X-axis starts at 0.5, the expected value for a classifier which assigns labels randomly, and AUCs less than 0.5 are not shown. ROC curves for all datasets are in Supplementary Figure \ref{fig:roc_curves}. Note that Youngster et al. (2014) \cite{cdi-youngster} had only 4 distinct control patients was excluded from the Random Forest analysis. (B) Left: Number of genera with q $<$ 0.05 (Kruskal-Wallis (KW) test, Benjamini-Hochberg FDR correction) for each dataset. If a study has no significant associations, no point is shown. Right: Direction of the microbiome shift, i.e. the percent of total associated genera which were enriched in diseased patients. In datasets on the leftmost blue line, 100\% of associated (q $<$ 0.05, FDR KW test) genera are health-associated (i.e. depleted in patients relative to controls). In datasets on the rightmost red line, 100\% of associated (q $<$ 0.05, FDR KW test) genera are disease-associated (i.e. enriched in patients relative to controls). Supplementary Figures \ref{fig:overall_heatmap_qvalues} and \ref{fig:overall_heatmap_foldchange} show q-values and effects for each genus in each study.}
	\label{fig:fig1}
	\end{center}
\end{figure}

\newpage
\begin{figure}[h]
	\begin{center}
  	\makebox[\textwidth][c]{\includegraphics[width=0.9\textwidth]{disease_specific_heatmaps.pdf}}
    \captionsetup{font=footnotesize,labelfont=footnotesize}
	\caption{\textbf{Comparing results from multiple studies of the same disease reveals patterns in disease-associated microbiome alterations.} Heatmaps showing log10(q-values) for each disease (Kruskal-Wallis (KW) test, Benjamini-Hochberg FDR correction). Rows include all genera which were significant in at least one dataset within each disease, columns are datasets. Q-values are colored by direction of the effect, where red indicates higher mean abundance in disease patients and blue indicates higher mean abundance in controls. Opacity ranges from q = 0.05 to 1, where q values less than 0.05 are the most opaque and q values close to 1 are gray. White indicates that the genus was not present in that dataset. Within each heatmap, rows are ordered from most disease-associated (top) to most health-associated (bottom) (i.e. by the sum across rows of the log10(q-values), signed according to directionality of the effect). The extent of a disease-associated microbiome shift can be visualized by the number of rows in each disease heatmap; the directionality of a shift can be seen in the ratio of red rows to blue rows within each disease. See Supplementary Figure \ref{fig:supp_dis_specific} for genus (row) labels.}
	\label{fig:fig2}
	\end{center}
\end{figure}

\newpage
\begin{figure}[h]
	\begin{center}
  	\makebox[\textwidth][c]{\includegraphics[width=\textwidth]{core_response.pdf}}
	\caption{\textbf{The majority of disease-associated microbiome alterations overlap with a non-specific microbial response to disease.} (A) Non-specific and disease-associated genera. Genera are in columns, arranged phylogenetically according to a PhyloT tree built from genus-level NCBI IDs (\url{http://phylot.biobyte.de}). Non-specific genera are associated with health (or disease) in at least two different \textit{diseases} (q $<$ 0.05, Kruskal-Wallis (KW) test, Benjamini-Hochberg FDR correction). Disease-specific genera are significant in the same direction in at least two \textit{studies} of the same disease (q $<$ 0.05, FDR KW test). As in Figure \ref{fig:fig2}, blue indicates higher mean abundance in controls and red indicates higher mean abundance in patients. Black bars indicate mixed genera which were associated with health in two diseases and also associated with disease in two diseases. Disease-specific genera are shown for diseases with at least 3 studies. Phyla, left to right: Euryarchaeota (brown), Verrucomicrobia Subdivision 5 (gray), Candidatus Saccharibacteria (gray), Bacteroidetes (blue), Proteobacteria (red), Synergistetes (pink), Actinobacteria (green), Firmicutes (purple), Verrucomicrobia (gray), Lentisphaerae (pink), Fusobacteria (orange). See Supplementary Figure \ref{fig:supp_meta_heatmap} for genus labels. (B) The percent of each study's genus-level associations which overlap with the shared response  (q $<$ 0.05, FDR KW test). Only datasets with at least one significant association are shown. (C) Overall abundance and ubiquity of non-specific genera across all patients in all datasets. Non-specific genera on the x-axis are as defined above.}
	\label{fig:fig3}
	\end{center}
\end{figure}

\bibliographystyle{unsrt}
\bibliography{meta-analysis/refs}
